% -*- mode: latex; tab-width: 2; indent-tabs-mode: nil; -*-
%------------------------------------------------------------------------------
% MRISC32 ISA Manual - Application Binary Interface.
%
% This work is licensed under the Creative Commons Attribution-ShareAlike 4.0
% International License. To view a copy of this license, visit
% http://creativecommons.org/licenses/by-sa/4.0/ or send a letter to
% Creative Commons, PO Box 1866, Mountain View, CA 94042, USA.
%------------------------------------------------------------------------------

\chapter{Application Binary Interface}
\label{sec:abi}

This chapter contains recommendations for platform application binary
interfaces (ABIs). It is not a complete ABI specification.

\section{Calling convention}

\subsection{Scalar registers}

\begin{tabular}{|c|c|l|}
  \hline
  \textbf{Register} & \textbf{Alias} & \textbf{Role and rule} \\
  \hline
  Z & R0 & Always zero (read only) \\
  \hline
  R1-R8 & & Function arguments / results \\
  \hline
  R9-R14 & & Temporary registers \\
  \hline
  R15 & & Intra-procedure call scratch register / temporary register \\
  \hline
  R16-R26 & & Callee-saved registers \\
  \hline
  TP & R27 & Thread pointer (callee-saved) \\
  \hline
  FP & R28 & Frame pointer (callee-saved) \\
  \hline
  SP & R29 & Stack pointer (callee-saved) \\
  \hline
  LR & R30 & Link register (callee-saved) \\
  \hline
  VL & R31 & Vector length (callee-saved) \\
  \hline
\end{tabular}

\subsubsection{Function arguments / results}

The first arguments to a function are passed in registers R1 to R8. How many
registers are used depends on the number of arguments and the types of the
arguments. For more information, see \ref{sec:abi_function_arguments}.

Likewise function results are returned in R1 to R8. For more information, see
\ref{sec:abi_function_results}.

These registers may also be used as temporary registers.

\subsubsection{Temporary registers}

Temporary registers are not guaranteed to be preserved across function call
boundaries, and thus need not be preserved by the callee.

\subsubsection{Callee-saved registers}

The contents of callee-saved registers must be preserved by a function. This is
normally done by the function prologue and epilogue by storing and restoring
the registers to and from the stack.

\subsubsection{Intra-procedure call scratch register}

The intra-procedure call scratch register may be used for call target address
calculations. It may also be used as a temporary register.

\subsubsection{Thread pointer}

The thread pointer may be used by systems that need to provide fast access to
thread local data. Otherwise it may be used as a general purpose register.

The thread pointer is a callee-saved register.

\subsubsection{Frame pointer}

\tbd

\subsubsection{Stack pointer}

Upon function entry, the stack pointer contains the address of the top of the
stack. For more information, see \ref{sec:abi_stack}.

The stack pointer is a callee-saved register.

\subsubsection{Link register}

The link register contains the return address to the caller.

The link register is a callee-saved register.

\subsubsection{Vector length}

The vector length is a callee-saved register.

\subsection{Vector registers}

\begin{tabular}{|c|c|l|}
  \hline
  \textbf{Register} & \textbf{Alias} & \textbf{Role and rule} \\
  \hline
  VZ & V0 & Always zero (read only) \\
  \hline
  V1-V8 & & Function arguments / results \\
  \hline
  V9-V15 & & Temporary registers \\
  \hline
  V16-V31 & & Callee-saved registers \\
  \hline
\end{tabular}

\todo{Do we really need that many callee-saved vector registers?}

\subsection{Stack}
\label{sec:abi_stack}

\tbd

\subsection{Function arguments}
\label{sec:abi_function_arguments}

\tbd

\subsection{Function results}
\label{sec:abi_function_results}

\tbd

\section{Data organization}

\subsection{Endianness}

Data fields are stored in memory using little endian representation. Thus the
least significant byte of a data field is at the lowest byte address that the
data field occupies in memory.

\subsection{Alignment}

Data fields that are one, two or four bytes in size shall be aligned to a
memory address that is divisable by the data field size.

Data fields that are larger than four bytes in size shall be aligned to a
memory address that is divisable by four.

\begin{tabular}{|l|l|l|}
  \hline
  \textbf{Type} & \textbf{Size (bytes)} & \textbf{Alignment (bytes)} \\
  \hline
  byte & 1 & 1 \\
  \hline
  half-word & 2 & 2 \\
  \hline
  word & 4 & 4 \\
  \hline
  double-word & 8 & 4 \\
  \hline
  quad-word & 16 & 4 \\
  \hline
\end{tabular}
